---
title: "Analysis of CONN Collection"
author: "James Mickley"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
graphics: yes
output:
  github_document:
    toc: yes
  html_document:
    keep_md: yes
    theme: readable
editor_options:
  chunk_output_type: console
---

```{r setup, include = F}
# @@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@
# @@@@@ Knitr Options
# @@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@

# Set root directory to the project directory
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())


# Set default knitr options: 
# Suppress warnings and messages, cache chunks, 
#  set default figure size to 6x8 at 300 dpi, and save a png and pdf
knitr::opts_chunk$set(warning = F, message = F, collapse = T, cache = T,
    fig.height = 6, fig.width = 8, dpi = 300, # 6x8" @ 300dpi:1800x2400=4.3MP
    dev = c('png', 'pdf'), dev.args = list(pdf = list(onefile = F)))
getwd()
```



## Overview

A quick analysis of the CONN collection, based on data from the [NEHerbaria Portal](http://neherbaria.org/).  I pulled all of the records from CT (205,000), of which 70,000 were from CONN.  

For now, restricting the data to CONN, though this could be run on the full dataset as well.

**Note:** The locality-based portion of this analysis does not include location-restricted records, so rare species are likely underrepresented. I don't have this data.

### Summary of Results
* 

```{r Main_Code, include = F, cache = F}

# @@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@
# @@@@@ Setup - This code is run, but output is hidden
# @@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@

# Load Packages
library(tidyverse) # Needed for data wrangling: dplyr, tidyr, ggplot2
library(cowplot) # Needed for publication-quality ggplots
library(knitr) # Needed to print tables with kable()
library(lubridate) # Needed for date functions
library(ggmap) # Needed for google maps
library(sf)

# Register James's Google API key for making google maps with ggmap
register_google(key = "AIzaSyC4ktT5QUVfEpHIUQCDQ2GN9_lzJm9M4FM")

# @@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@
# @@@@@ Data Preparation
# @@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@

# Import datasets
data <- read.csv("data/occurrences.csv")

# @@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@
# @@@@@ ggPlot Theme
# @@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@

ggplot.theme <- theme(
    
    # Text size for axis ticks
    axis.text.y = element_text(size = 14),
    axis.text.x = element_text(size = 14),
    
    # Text size for axis labels
    # Also move them away from the axes a bit for more space
    axis.title.x = element_text(size = 18, face = "bold", vjust = -1),
    axis.title.y = element_text(size = 18, face = "bold", vjust = 1.5),
    
    # Plot title size
    plot.title = element_text(size = 20, face = "bold"),
    
    # Margins for top, right, bottom, left
    plot.margin = grid::unit(c(1.5, 1.5, 1.5, 1.2), "lines"), 
    
    # Legend text size
    legend.text = element_text(size = 14),
    legend.text.align = 0, 
    legend.title = element_text(size = 16, face = "bold"),
    legend.key.size = grid::unit(1.4, "line"),
    legend.key = element_blank()
    )

# Read in connecticut counties shapefile
ct.shp <- st_read("data/shapefiles/CTCounties/wgs84/countyct_37800_0000_2010_s100_census_1_shp_wgs84.shp")

# Construct a google map of connecticut to use as a base
ct.gmap <- qmap(location = "Connecticut", zoom = 8, maptype = "hybrid") +
    
    # Add connecticut counties
    geom_sf(data = ct.shp, inherit.aes = FALSE, fill = NA, color = "blue",
        size = 1) + 
    
    # Set a bounding box
    xlim(-73.8, -71.7) + ylim(40.95, 42.05) + 
    
    # Adjust the map theme
    theme(
        axis.text = element_blank(),
        axis.line = element_blank(),
        axis.ticks = element_blank(),
        panel.border = element_blank(),
        axis.title = element_blank(),
        panel.grid.major = element_line(colour = "transparent"), 
        panel.grid.minor = element_line(colour = "transparent"), 
        panel.background = element_rect(fill = "transparent", colour = NA)
    )


```


```{r Data_Wrangling, include = F}



# Wrangle the main dataset, restricting to CONN, and cleaning it up
conn <- data %>% 
    
    # Restrict to only collections from the CONN herbarium
    filter(collectionCode == "CONN") %>%
    
    # Remove genus level specimens
    filter(specificEpithet != "") %>%
    
    # Remove years that are zero
    mutate(year = ifelse(year == 0, NA, year)) %>%
    
    # Make a date column from year, month day, 
    # setting to 1/1 for dates missing month and day
    mutate(date = ymd(ifelse(is.na(year), NA, 
        paste(year, ifelse(is.na(month), 1, as.numeric(month)), 
            ifelse(is.na(day), 1, as.numeric(day)), sep = "-"))))
    

# Calculate the date for which each species was last collected in CT
ct.lastcollected <- conn %>% 
    
    # Remove dates without a year
    filter(!is.na(year)) %>%
    
    # Group by species
    group_by(scientificName) %>%
    
    # Count the number of specimens of each species
    mutate(specimens = n()) %>%
    
    # Ungroup
    ungroup() %>%

    # Arrange by date in descending order
    arrange(desc(date)) %>%

    # Only keep the row for the most recent collection date for each species
    distinct(scientificName, .keep_all = T) %>%
    
    # Select columns to keep
    select(order, family, scientificName, genus, specificEpithet, taxonRank, 
        infraspecificEpithet, lastcollected = year, lastdate = date) %>% 
    
    # Convert to dataframe
    data.frame()


# Calculate the date for which each species was last collected by county
county.lastcollected <- conn %>% 
    
    # Remove records without a defined county
    filter(county != "Middlesex/New Haven", county != "") %>%
    
    # Remove dates without a year
    filter(!is.na(year)) %>%
    
    # Group by county and species species
    group_by(county, scientificName) %>%
    
    # Count the number of specimens of each species by county
    mutate(specimens = n()) %>%
    
    # Ungroup
    ungroup() %>%

    # Arrange by date in descending order
    arrange(desc(date)) %>%

    # Only keep the row for the most recent collection date for each species/county
    distinct(county, scientificName, .keep_all = T) %>%
    
    # Select columns to keep
    select(county, order, family, scientificName, genus, specificEpithet, taxonRank, 
        infraspecificEpithet, lastcollected = year, lastdate = date) %>% 
    
    # Convert to dataframe
    data.frame()


# Cutoff for the number of years since a species was collected
threshold = 50

# Make a dataset of all the records belonging to species that haven't been 
# seen in the threshold number of years across the state
ct.old.collections <- conn %>% 
    
    # Merge in last collected date
    left_join(select(ct.lastcollected, scientificName, lastcollected), 
        by = "scientificName") %>%
    
    # Filter to only records before the cutoff threshold # of years
    filter(lastcollected < year(now()) - threshold)


# Write data out to csv
write.csv(ct.old.collections, "data/old-collections/ct-specimens.csv",
    row.names = F)


# Make a dataset of all the records belonging to species that haven't been 
# seen in the threshold number of years for each county
county.old.collections <- conn %>% 
    
    # Merge in last collected date
    left_join(select(county.lastcollected, county, scientificName, lastcollected),
        by = c("county", "scientificName")) %>%
    
    # Filter to only records before the cutoff threshold # of years
    filter(lastcollected < year(now()) - threshold)

# Write data out to csv
write.csv(county.old.collections, "data/old-collections/county-specimens.csv",
    row.names = F)


```




## Sampling

```{r Species_Lists, include = F}


# Compile a list of species for the state from CONN data
ct.spplist <- conn %>% 
    
    # Group by species, keeping higher ranks
    group_by(order, family, genus, scientificName) %>% 
    
    # Count the number of specimens of each species
    summarize(specimens = n()) %>% 
    
    # Ungroup the dataset
    ungroup() %>%
    
    # Merge in last collected date for the state
    left_join(select(ct.lastcollected, scientificName, lastcollected), 
        by = "scientificName") %>%
    
    # Arrange species from most to least common
    arrange(desc(specimens))


# Compile a list of species by county from CONN data
county.spplist <- conn %>% 
    
    # Group by county and species, keeping higher ranks
    group_by(county, order, family, genus, scientificName) %>% 
    
    # Count the number of specimens of each species by county
    summarize(specimens = n()) %>% 
    
    # Ungroup the dataset
    ungroup() %>%
    
    # Merge in last collected date for each county
    left_join(select(county.lastcollected, county, scientificName, 
        lastcollected), by = c("county", "scientificName")) %>%
    
    # Arrange species from most to least common within each county
    arrange(county, desc(specimens))


# Write out overall state species list
write.csv(ct.spplist, "data/species-lists/ct-specieslist.csv", row.names = F)

# Write out species lists for each county:

# Fairfield County
write.csv(county.spplist %>% filter(county == "Fairfield"),
    "data/species-lists/fairfield-specieslist.csv", row.names = F)

# Hartford County
write.csv(county.spplist %>% filter(county == "Hartford"),
    "data/species-lists/hartford-specieslist.csv", row.names = F)

# Litchfield County
write.csv(county.spplist %>% filter(county == "Litchfield"),
    "data/species-lists/litchfield-specieslist.csv", row.names = F)

# Middlesex County
write.csv(county.spplist %>% filter(county == "Middlesex"),
    "data/species-lists/middlesex-specieslist.csv", row.names = F)

# New Haven County
write.csv(county.spplist %>% filter(county == "New Haven"),
    "data/species-lists/newhaven-specieslist.csv", row.names = F)

# New London County
write.csv(county.spplist %>% filter(county == "New London"),
    "data/species-lists/newlondon-specieslist.csv", row.names = F)

# Tolland County
write.csv(county.spplist %>% filter(county == "Tolland"),
    "data/species-lists/tolland-specieslist.csv", row.names = F)

# Windham County
write.csv(county.spplist %>% filter(county == "Windham"),
    "data/species-lists/windham-specieslist.csv", row.names = F)



```


### Statewide

```{r State_List, echo = F}

# Make a summary by county: families, species, and # specimens
ct.spplist %>% 
    
    # Group by both county and family
    group_by(family) %>% 
    
    # Count up the number of species and specimens by family for each county
    summarize(species = n(), specimens = sum(specimens)) %>%
    
    # Summarize # of families, species, specimens for CT
    summarize(families = n(), species = sum(species), 
        specimens = sum(specimens)) %>%
    
    # Output a table of the data
    kable(format = "markdown")

```

A statewide species list, with # of specimens per species and last collected year is available at [ct-specieslist.csv](data/species-lists/ct-specieslist.csv).

The northeast corner of the state (Windham County), nothern New London County, and the southern half of Litchfield County look more sparsely sampled.

```{r State_Timeline, echo = F}

# Plot a histogram of specimens across time
ggplot(conn, aes(x = year)) + 
    
    # Add histogram with 1 bin per year
    geom_histogram(bins = 200, fill = "blue") + 
    
    # Set x axis limits (200 years)
    xlim(1820, 2020) + 
    
    # Add a title
    ggtitle("Sampling by Year")

```


```{r State_Map, echo = F}

# Plot all the records with location information on a google map
ct.gmap + 
    
    # Add points for each record
    geom_point(data = conn, aes(x = decimalLongitude, y = decimalLatitude), 
        color = "red", size = 1.2, alpha = 0.3) + 
    
    # Add title
    ggtitle("All CONN Records")

```



### By County

Windham and Middlesex counties stand out as having less sampling, both across time and overall. They were largely ignored prior to 1970. Litchfield, New London and Fairfield, and Tolland have been comparatively well sampled recently, slightly less so for New Haven.

County level species lists, with # of specimens and last collected year for that county are available in [species-lists](data/species-lists/).


```{r County_List, echo = F}

# Make a summary by county: families, species, and # specimens
county.spplist %>% 
    
    # Remove records without a defined county
    filter(county != "Middlesex/New Haven", county != "") %>%
    
    # Group by both county and family
    group_by(county, family) %>% 
    
    # Count up the number of species and specimens by family for each county
    summarize(species = n(), specimens = sum(specimens)) %>%
    
    # Group by county
    group_by(county) %>%
    
    # Summarize families, species, and specimens by county
    summarize(families = n(), species = sum(species), 
        specimens = sum(specimens)) %>% 
    
    # Output a table of the data
    kable(format = "markdown")


```

```{r County_Timeline, echo = F}

# Plot a histogram of specimens across time
conn %>% 
    
    # Remove records without a defined county
    filter(county != "Middlesex/New Haven", county != "") %>%
    
    # Make plot
    ggplot(aes(x = year)) + 
    
        # Make a separate graph facet for each county
        facet_wrap(~ county) + 
    
        # Add histogram with 1 bin per year
        geom_histogram(bins = 200, fill = "blue") + 
        
        # Set x axis limits
        xlim(1820, 2020) + 
        
        # Add a title
        ggtitle("Sampling by Year") + 
    
        ggplot.theme

```


### By Taxonomic Order

Less useful than I thought, leaving it up just in case.  The undersampled species are widely distributed taxonomically I guess.

```{r Order_Sampling, echo = F, fig.width = 6, fig.height = 16}
     
ct.spplist %>% filter(order != "") %>%
    
    # Reverse the factor levels of order so that the graph starts with A
    mutate(order = factor(order, levels = rev(levels(order)))) %>%

    # Make a plot showing the spread of # of specimens by order
    ggplot(aes(x = order, y = specimens)) +
        
        # Add points for each species
        geom_jitter(width = 0.1, height = 0, alpha = 0.5, color = "tomato") +
        
        # Add transparent boxplot
        geom_boxplot(alpha = 0) + 
        
        # Make the y-axis on a log scale, with the following tickmarks
        scale_y_log10(position = "right") +
        
        # Flip the x and y axes for better readability
        coord_flip() +
        
        # Add axis labels
        ylab("specimens per Species by Order") + xlab("Order") + 
        
        # Black and white theme
        theme_bw()

```

## Undersampled Species

One way to get at data-deficient species, is to look at the number of species with very few records.

Of the 4600 species recorded, more than half have less than 10 specimens statewide, and over 1000 species are only represented by one observation. These would be a good target, though it's possible some are outdated names or something of that sort.

Notably though, most of the undersampled species were collected recently. There are pockets of undersampled species that haven't been collected recently in the CT River Valley, New London, and the NW corner that might be worth looking into.

```{r Undersampled_Species, echo = F}

# Filter to only include species with 10 or fewer specimens
ct.spplist %>% filter(specimens <= 10) %>% 
    
    # Group by the number of specimens
    group_by(specimens) %>% 
    
    # Summarize the number of species with a certain number of specimens
    summarize(species = n()) %>% 
    
    # Reorder and rename columns
    select(species, num.specimens = specimens) %>%
    
    # Output a table of the data
    kable(format = "markdown")


```


```{r Undersampled_Map, echo = F}

# Make a dataframe of all species records that belong to an undersampled species
undersampled <- ct.spplist %>% filter(specimens <= 10) %>% 
    
    # Join the records in conn (with lat/long)
    inner_join(conn, by = "scientificName") 


# Map the undersampled species    
ct.gmap + 
    
    # Add points for each record
    geom_point(data = undersampled, 
        aes(x = decimalLongitude, y = decimalLatitude, color = lastcollected), 
        size = 2, alpha = 1) + 
        
    # Color by a rainbow gradient
    scale_color_gradientn(name = "Last Collected", colors = rainbow(6)) + 
    
    # Add title
    ggtitle("Undersampled Species")
    
```


## Number of Species Not Collected in 50 Years

A second way of looking for data-deficient species is finding species that haven't been collected in a while. These species could be extinct in CT (especially for prairie species), but they could also be difficult to identify (Poales), rare, or infrequently collected.

I set the threshold at 50 years or 1969.

### For The State

```{r, Old_Species, echo = F}

# Summarize the total number of species for each family
families <- ct.spplist %>% group_by(family) %>% 
    summarize(total = n())

# Summarize the # of species not recently collected by family
ct.old.species <- ct.old.collections %>% 

    # Group by scientific name, keeping other columns
    group_by(order, family, genus, scientificName, lastcollected) %>%
    
    # Summarize to one row per species
    summarize() %>% 
    
    # Remove the grouping
    ungroup() %>%

    # Remove blank families
    filter(family != "") %>%
    
    # Group by family
    group_by(family) %>%
    
    # Count the number of species in each family not recently collected
    summarize(species = n()) %>%
    
    # Add the total number of species per family
    inner_join(families, by = "family") %>%
    
    # Calculate the proportion of species in a family not recently collected
    mutate(proportion = 1 - (species/total))
    
    
```

For all of CT, the families with the most of these sorts of species are plotted below.  Poaceae and Cyperaceae are well-represented, but there is also a lot of missing diversity in common families: Asteraceae, Rosaceae, Brassicaceae, Lamiaceae, Fabaceae, etc.

Another way to look at this is by looking at the proportion of species in the family that haven't been collected recently. This gives a different set of families. Both would make good collecting targets.


A static map of where these species are is below. Many were collected in New London, Storrs, and the CT River Valley.

Or there's an [interactive map](http://mickley.github.io/CT-Herbarium-Specimens/analyses/CT-Old-Map.html) that lets you zoom in to particular areas and see which species were collected there.


```{r Old_State, echo = F, fig.width = 6, fig.height = 10}

# 
ct.old.species %>%
    
    # Restrict to families with 5 or more species not recently collected
    filter(species >= 5) %>%
    
    # Arrange orders by number of species not collected recently
    arrange(species, family) %>%
    
    # Reorder the factor levels to reflect the order above
    mutate(family = factor(family, unique(family))) %>% 

    # Make a plot showing the number of species not collected recently by order
    ggplot(aes(x = species, y = family)) +
        
        # Add scatterplot points
        geom_point() + 
        
        # Add axis labels
        xlab("# Species Not Collected Since 1969") + 
        ylab("Families With 5+ Species") + 
        
        # Black and white theme
        theme_bw()


```



```{r, Old_State_Prop, echo = F}

# Graph the proporion of species in each family that have not been 
# collected recently
ct.old.species %>%
    
    # Filter to families with more than 5 species, with less than 60% recent
    filter(total >= 5, proportion < 0.6) %>% 
    
    # Arrange by descending proportion
    arrange(desc(proportion)) %>% 
    
    # Reorder the factor levels to reflect the order above
    mutate(family = factor(family, unique(family))) %>% 
    
    # Plot
    ggplot(aes(x = proportion, y = family)) +
        
        # Add scatterplot points
        geom_point() + 
        
        # Add axis labels
        xlab("Proportion of Species Collected Since 1969") + 
        ylab("Families With 5 + Species") + 
        
        # Black and white theme
        theme_bw() + ggplot.theme


```



```{r Old_State_Map, echo = F}

# Map the species 
ct.gmap + 
    
    # Add points for reach record
    geom_point(data = ct.old.collections, 
        aes(x = decimalLongitude, y = decimalLatitude), 
        color = "red", size = 2, alpha = 0.3) + 
    
    # Color by a rainbow gradient
    scale_color_gradientn(name = "Last Collected", colors = rainbow(6)) + 
    
    # Add title
    ggtitle("Species Not Collected Since 1969")


```




```{r Old_County, echo = F, eval = F}



# 
county.old.species <- county.old.collections %>% 

    group_by(county, order, family, genus, scientificName, lastcollected) %>%
    
    summarize() %>% 
    
    ungroup()


# 
county.old.species %>%
    
    group_by(county) %>%
    
    summarize(species = n()) %>%
    
    # Arrange orders by number of species not collected recently
    arrange(species) %>%
    
    # Reorder the factor levels to reflect the order above
    mutate(county = factor(county, unique(county))) %>% 

    # Make a plot showing the number of species not collected recently by order
    ggplot(aes(x = species, y = county)) +
        
        # Add scatterplot points
        geom_point() + 
        
        # Add axis labels
        xlab(paste("# Species Not Collected Since", year(now()) - threshold)) + 
        ylab("County") + 
        
        # Black and white theme
        theme_bw()



```



## Session Information

```{r Session_Info, echo = F, comment = ""}

# Add session information to help with reproduceability
sessionInfo()


```


